Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-15 VegOzone.tex
Check intro duction with grammarly

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone causes reduced photosynthesis, visible foliar injury, early senescence and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality \citep{NCC:Lin2020}. Tropospheric background ozone concentrations derived from satellite instruments range between $50-65\,\unit{ppb}$ over Fennoscandia but strongly depend on season \citep{ESA:Cooper2014}.

@@

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway experienced temperatures $4\,\unit{^\circ C}$ above normal and saw only about $60\,\unit{\%}$ of normal precipitation. Northern Norway as a whole had close to normal precipitation, one of the causes of the high rate of forest fires in that area.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors.
A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Boreal wildfires are found to emit ozone precursors \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}. Coincident peak \chem{[O_3]} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018 shall serve as a reference for probable future conditions in northern Fennoscandia. Conditions in 2019 represent present climate.\\

>>>>

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone causes reduced photosynthesis, visible foliar injury, early senescence, and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions concerning ozone impacts and air quality \citep{NCC:Lin2020}. Tropospheric background ozone concentrations derived from satellite instruments range between $50-65\,\unit{ppb}$ over Fennoscandia but strongly depend on season \citep{ESA:Cooper2014}.

@@

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway experienced temperatures $4\,\unit{^\circ C}$ above normal and saw only about $60\,\unit{\%}$ of normal precipitation. Northern Norway as a whole had close to normal precipitation, one of the causes of the high rate of forest fires in that area.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors.
A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Boreal wildfires are found to emit ozone precursors \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}. Coincident peak \chem{[O_3]} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018 shall serve as a reference for probable future conditions in northern Fennoscandia. Conditions in 2019 represent the present climate.\\
